### 2009-12-12 docs/marcello-dissertation/thesis/chapter8-experimental-results-correct-behavior-and-performance.tex
Revisions from Niki on Chapters 7 and 8... (artifacts 8 and 9 because 7 was deleted)
Grammatical errors, and addition of a section to chapter 7.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@698 559edde2-a4d4-44e0-96fc-4035e40d7396

As discussed in the previous chapter, the implementation of a data persistence
layer for NetBEAMS followed the requirements defined in chapter 4 having in
mind the classification of the SF-BEAMS sensor network and the software
infrastructure provided by the Data Sensor Platform. As a result, the
implementation was developed using the infrastructure provided by the online
version-control system repository from NetBEAMS, whose documentation is
detailed in section \ref{sec:dsp-data-persistence-implementation}.

First, this chapter details the experiment setup designed to evaluate the
proposed implementation, showing the measurement results collected from the
simulation log and based on the user experience while reusing the collected
data. Last, but not least, the pertinent discussion of this work is devided up
into different sections, each of them analyzing different aspects related to
data persistence for NetBEAMS, amongo others.

@@

The experiment was designed to support different scenarios of a regular use of
the mongoDB system infrastructure, simulating the activities of data collection
from the SF-BEAMS network using the NetBEAMS infrastructure. First, the setup
for the data collection of randomly generated data was implemented, which
directly uses NetBEAMS already existing classes. Then, the implementation of
the CRUD scenarios were implemented in different programming languages, from
the data insertion scenario using the Java Programming Language, and then the
remaining operations using the Javascript Scripting language \cite{javascript}
and the Python Programming Language \cite{python}. Similarly, an open-source web
application, developed in the Ruby Programming language \cite{ruby}, was also
used. The experiment scripts were writen using Shell Bash Script Language
\cite{bashshell} and is responsible for orchestrasting the execution of one
single experiment ``run''.

@@

Different artifacts were created to perform the different experiments
operations. The Create function was developed in Java, without the interversion
of the DSP or OSGi, isolating only the functionalities of provided by CRUD
service class DSPMongoCRUDService (Listing \ref{file:dsp-mongo-service}),
responsible for the connection handling with mongoDB, as well as the insertion
operation from NetBEAMS. As described in section \ref{sec:dsp-details}, the
current version of NetBEAMS supports the data collection for the YSI Sonde
device, among others used for demo purposes such as a software sensor that
captures the mouse activities over a window. The documentation of both data
types is seen in section \ref{sec:dsp-payload-implementation}. In this way, the
DSP Data Persistence, as described in the previous section, supports both data
types to be persisted.

In order to exercise mongoDB with the same data collected from NetBEAMS, the
existing Java class TestSondeData (Listing
\ref{file:random-ysi-data-generator}), was refactored to provide a random
instance generator for the YSI Data handler implemented in the Java class
SondeDataType (Listing \ref{file:main-ysi-data-handler}). When the Insert
function is executed, the created instances are transferred from the JVM to
the mongoDB, and stored in the collection ``SondeDataContainer''. Specific
details about the execution of mongoDB are depicted in section
\ref{sec:mongodb-deployment}. Finally, the implementation of the experiment
in the Java class DSPMessageToMongoDBExperiment (Listing
\ref{file:experiment-dsp-java-executor}) was responsible for managing the
creation and insertion of any workload composed by instances of the class
SondeDataType in a bulk manner.

@@

On the other hand, the execution of the main experiment shell script requires
the use to provide the size of the workload, as shown in listing
\ref{cmd:run-persistence-experiment}.

\lstset{label=cmd:run-persistence-experiment,caption=Running main experiment
shell script}
\begin{lstlisting}
marcello@netbeams-dev:~/workspaces/netbeams/versions/v2/persistence$ ./run-persistence-experiment 
#######  NetBEAMS Experiments - Persistence on MongoDB  ########

Usage: ./run-persistence-experiment X, where X is the size of the workload to be inserted from NetBEAMS YSI Data Handler to mongoDB
\end{lstlisting}

@@

The workload selected for the experiments reflects the current use of the data
at the SF-BEAMS infrastructure. In this way, the experiment runs for 5
different times performed to investigate different scenarios to be described.
In this way, this number is related to the total number of sensor devices
reported to be in operation, as discussed in section \ref{sec:sfbeams}. In
summary, the volume of the workload prepared after the experiment setup
is as follows:

\begin{itemize}
  \item \textbf{First Round}: 483,840 documents, worth the production of the 
  YSI Sonde device during one year, at the rate of 1 observation per minute;
  \item \textbf{Next Rounds}: 483,840 * 4, or 2,419,200 documents, representing
  5 different YSI devices similar to the first round.
\end{itemize}

@@

Two different system environments were prepared: one for the External Storage
environment, or Single Server; and another for the Data-Centric Storage, or
Distributed Server. Since the single server is a regular centralized database
system, mongoDB just requires the execution of the database process ``mongod'',
which is started by specifying the directory where the inserted data will be
managed. On the other hand, nonetheless, the distributed server setup requires
a different list of processes (see section \ref{sec:mongodb-user-experience}):
one ``mongod'' process for each mongoDB shard running on the different mongoDB
cluster, one ``mongod'' process to manage the cluster metadata, and one
``mongos'' process, which is the main one responsible to orchestrate the
cluster. In conclusion, the single system environment is synmmetric to
the External Storage used to persist the collected data, while the clustered one
is simulates the Data-Centric Storage one.

@@

In order to verify the feasibility of the solution proposed, all CRUD
operations were exercised through the implementation of the practical
scenarios based on the use cases described in section \ref{sec:use-cases}. In
brief, the use cases were chosen to run with random values such as follows:

\begin{itemize}
  \item (R0) Insert a volume of collected data from sensors using NetBEAMS
  service compatible with 1 year of observations;
  \item (R1) Find all documents whose observations were collected between two
  different dates, say the days between December 8 and 15 of 2009;
  \item (R2) Find all documents whose observations were collected from a given
  known sensor device such as one whose IP address is equals to "192.168.0.102";
  \item (R3) Find all documents whose observations contained specific values
  read from the environment, say Salinity equals to 0.01 and the water temperature
  equals to 46.47;
  \item (U1) Update all the documents whose observations were collected during
  the day of December 2nd, 2009, by adding a tag = ``oil spill'';
  \item (D1) Remove all the documents whose observations were collected on
  December 7th, 2009.
  \item (R4) Export the produced data of the week to CSV files, using the same
  tabular format sequence used in the files distributed by SF-BEAMS using the 
  the OPEnDAP format, as in Listing \ref{file:rtc-ysi-opendap}.
\end{itemize}

@@

As described in the previous section, the design of the experiments were based
on the specifications of a data persistence for NetBEAMS, using the mongoDB
database system. In this way, after setting up the experiment environment,
the execution of the workload was performed against the mongoDB database using
its mechanisms used to manipulate the data generated. It uses the programming
language abstraction to provide access and modifiy the data collected from the
NetBEAMS component. For this reason, this section details these mechanisms uses
to implement each of the scenarios defined in the previous section, associating
numerical response time collected from the logs.

@@

The first set of measurements collected were the ones regarding the use case
R1, shown in Listing \ref{file:dsp-mongo-service}, defined as the insertion of
instances of the collected data from the sensor devices using Java. The
document follows the specification of the keys designed in section
\ref{sec:dsp-persistence-data-model}. The use of the mongoDB Java driver, which
provides the methods to access the database system and the collection. In this
way, the construction of the keys and respective values were done using a
Hash-like object ``BasicDBObject'' such as the ones shown in the method
``buildKeySegment''. Similarly, the creation of database indexes is performed
using the same approach, as shown in the Java method ``setupMetadataIndexes''.
Finally, the insertion of the documents is performed on the service method
call ``insertPersistentUnitMessageContents'', which delimits the bulk insert
using the transaction management mechanism over the collection. In this way,
different insert times were collected from the logs of the different
experiments as summarized on Table \ref{tab:experiment-insert-avarage}.

\begin{table}
    \begin{center}
        \begin{tabular}{|p{100pt}|p{100pt}|p{100pt}|}\hline
           x & \textbf{Linux VirtualBox Server} & \textbf{Mac OS Host Server}\\\hline 
           \textbf{Single Server} & ~25K docs/sec & ~51K/sec \\\hline
           \textbf{Clustered Server} & ~17K docs/sec & ~39K/sec \\\hline
        \end{tabular}
        \caption{Insertion averages on Virtual and Host in Single or Clustered
        mongoDB Server}
    \end{center}
    \label{tab:experiment-insert-avarage}
\end{table}

After executing the experiments runs, the claimed disk space for
the representation of the randomly generated data is separated into two
different sizes: one related to the amount of data produced by one YSI device,
and another produced by five different YSI devices, as summarized on Table
\ref{tab:experiment-insert-diskspace}.

\begin{table}
    \begin{center}
        \begin{tabular}{|p{100pt}|p{100pt}|p{100pt}|}\hline
        x & \textbf{1 YSI} & \textbf{5 YSIs}\\\hline
        \textbf{Indexed Keys} & 278.33 MB & 1.35 GB \\\hline
        \end{tabular}
        \caption{Amount of disk storage used}
    \end{center}
    \label{tab:experiment-insert-diskspace}
\end{table}

@@

When it comes to the simulation of the use case scenarios, the implementation
used the mongoDB shell client, which uses Javascript, to implement most of the
operations for ``retrieve'', ``update'' and ``delete'' shown in Listing
\ref{file:experiment-scenarios}. On the other hand, the implementation of the
use case to export the collected data to OPeDAP format used Python, depicted on
Listing \ref{file:experiment-export-python}. The access pattern to use the
collected data is ``db.collection\underline{ }name.operation", where
``collection\underline{ }name'' is the name of the collection used for the
collected data, and in this case, the name of the DSP component that represents
the collected data. Therefore, mongoDB uses dynamic object binding for the
collection object, and executes the CRUD operations as method call with a list
of the needed parameters. Table \ref{tab:experiment-scenarios-mongo-methods}
relates method calls used to implement each of the use cases defined, as
implemented in Listing \ref{file:experiment-scenarios}.

\begin{table}
    \begin{center}
        \begin{tabular}{|c|c|c|c|}\hline
        \textbf{Scenario Type} & \textbf{Language} & \textbf{Method Call} & \textbf{Use Case}\\\hline 
        \textbf{Insert} & Java & db.SondeDataContainer.insert() & R0\\\hline
        \textbf{Retrive} & Javascript & db.SondeDataContainer.find() & R1, R2, R3\\\hline 
        \textbf{Update} & Javascript & db.SondeDataContainer.update() & U1\\\hline 
        \textbf{Delete} & Javascript & db.SondeDataContainer.remove() & D1\\\hline
        \textbf{Export} & Python & db.SondeDataContainer.find() & R4\\\hline
        \end{tabular}
        \caption{mongoDB method calls used for the use cases}
    \end{center}
    \label{tab:experiment-scenarios-mongo-methods}
\end{table}

The execution time for each of the use cases were collected using the logs
collected during the execution of the experiment script manager. The retrieval
operations ran in avarage for ~300ms, while the updates took more than 10
seconds when it involved larger data sets with more than 300 thousand
documents. Similarly, the delete operation took less

@@

In general, the use experience of mongoDB is related to the use of programming
languages, since most of the methods to create, access and modify the database
is performed using the programming language abstraction of objects and method
calls. The creation of data was the first contact with mongoDB in terms
of data creation in Java. Later, Different methods of data access were used
after the randomly generated data was inserted into mongoDB: using the shell
commands mongo and ``mongoexport'', command-line execution developed in
Python, through the REST Web Services API, and using a web application,
written in the Ruby Programming Language \cite{ruby}, providing the view over
the collected data using a web browser. More details about the execution of
these commands during the experiments is shown in section
\ref{sec:mongodb-user-experience}.

First and foremost, the primary method to access to the database is using the
shell command. Once the user requests an access to the database, he or she can
issue an operation over the collection of documents using the regular the
abstraction programming language, as the shown in Listing \ref{cmd:mongo}. The
CRUD operations are available to the user and can be used at any time to any
database collection. Aggregation operations such as to count the returned
resultset is also provided via programming languages abstraction.

In order to export the collected data to other data formats, the  command
``mongoexport'' and the developed export script written in Python were used.
The ``mongoexport'', exports the collected data from the database using the
native support to comma-delimited values. As shon in Listing
\ref{file:mongodb-export-command}, the actual execution, the remaining ones
describe how mongo is querying the data from the database. As a result, the
total number of documents exported is shown in on line 8. In addition to the
regular command that exports the entire database, the user can provide a query
string during the execution of the command in order to filter the dataset
returned and, consequently, the number of documents exported. In contrast to
the use of the default export capabilities of mongoDB, an Export script written
in Python, as shown in Listing \ref{file:experiment-export-python}. It uses the
package ``pymongo'' and creates a comma-delimited file with the same header
format and lines as shown in the HTTP Response body of Listing
\ref{file:rtc-ysi-opendap}. The results can be compared with the example of
Listing \ref{file:experiment-export-python-results}.

While the access to the database using the shell command gives full access to 
the CRUD operations, the access to the data was also performed using the REST
Web Services API, as shown in Listing \ref{cmd:mongo-rest-request}. As one can
see, the IP address and the default port number of the mongoDB REST middleware
are provided. Then, the REST operation uses the name of the database and
the collection to correlate with the ones in the database system. In this way,
querying the collections from mongoDB can be performed using the HTTP Request
Parameter variables, by adding the prefix ``filter\underline{ }'' before the
name of the key of the document. The given example shows the query performed in the
collected data for the key ``observation.Conductivity'' with the value
``104.5''. Similarly, the HTTP Request variable ``limit'' is used to reduce
the number of documents returned by the query. Finally, the HTTP Response from
mongoDB returns the result using the JSON format, similar to the ones printed
in the command-shell.

@@

The actual execution of the solution was performed using the demo Mouse
Actions, which consists of a software sensor that perceives any mouse action in
a window area. The experiment consisted of moving the mouse around a white
window, where each action is captured and temporarily saved until a threshold is
reached, making the component to send the messages to the persistence layer, so
that it can save on mongoDB. Figure \ref{fig:experiment-mouseactions-dsp} shows
a screenshot of the moment when the message is sent to mongoDB, being verified
through the HTTP REST Interface using a web browser.

@@

The implementation of the data persistence for NetBEAMS follows the
specifications explained in chapter 4. The purpose of sensor network for
NetBEAMS is to suport data archival from any collected data, and for this
reason, mongoDB was selected after an empirical analysis described in chapter
5. In this way, the requirements described in chapter 4 were expressed as
examples of use cases in section \ref{sec:exp-scenarios}. This chapter
discusses the experiments in different angles: analysis of the taxonomies, the
infrastructure and the use cases evaluation.

@@

Any database system covers the purpose of the sensor data for NetBEAMS, which
is Data Archival, along with the type of storage for the data as an External
Storage. After selecting mongoDB, the development of the persistence layer met
the requirements of archiving data for different types of sensor devices from
the DSP Platform such as the YSI Data Handler and the Mouse Actions demo data.
As it is shown in section \ref{sec:exp-measurements}, the collected data is
archived in files using the BSON format \cite{bson}, a binary representation
used to store instances of the document described in Listing
\ref{file:mongodb-ysi-data-format}. Along with the purpose of Data Archival,
the use of mongoDB also met the specification of an External Storage, since the
data persisted in the file system, as well as Data-Centric for running mongoDB
with different  distributed mongo shards (cluster nodes).

SF-BEAMS uses Tabular Data model, Schema-less data model that represents the
collected data using comma-delimited files, without a database system to manage
those files, just the OPEnDAP middleware. In contrast, the data persistence
proposed by this work proposes mongoDB as a schema-less database system to
manage the collected data collected using NetBEAMS. As described in chapter 5,
the schema-less data model can provide a better support to Data Provenance
because the data model gives support to the change of records without affecting
the others. In this way, the metadata used by the implementation clearly
describes the three fundamental questions regarding the collected data from
sensor devices, as described in section \ref{sec:dsp-persistence-data-model}.
First, the identity of the data is given by the key ``message\underline{
}id'', in a way that it uniquely identifies the data produced by the sensor
identified by the key ``sensor.ip\underline{ }address''. In this way, the
message can be tracked back to the sensor that produced the data for different
purposes such as tracking the data producer. Second, time dimensions were used
in order to track two different aspects of the data collection: the valid time
can be used to correlate observations to the exact moment it occurred, while
the transaction time can be used to verify when the data was collected, and
make decisions such as deleting the instances of data were collected in the
previous two days, etc. Last, but not least, the use of the key
``sensor.location'' can give the exact location of the data, since researchers
can be interested in finding data based on a specific location. Finally, the
most important aspect of the schema-less data model is that the observed data
is added to the key ``observation'', as any sensor device carries its set of
attributes of key-value pairs. A good example of the applicability of the
schema-less model is the scalability regarding changes to the data structure.
For instance, if an existing sensor device is flushed with a new version of
the firmware\footnote{Primary machine instructions that run in the hardware},
changing the format of data types or adding new values, the following instances
of collected data may be different to the previous one by simply adding the
new additions of data. For this reason, the DSP Data Persistence gives
administrators the ability to choose which fields to persist by declaring it
using a bootstrap message for the component, as shown in the parameter
``YSI\underline{ }DESIRED\underline{ }PROPERTIES'' of Listing
\ref{file:dsp-data-persistence-bootstrap.xml}.

In order to access the collected data, the implementation used different APIs
provided by the mongoDB community. The Centralized Query Mechanism was
satisfied by the use of those APIs, which is also fultill the requirements
defined by NetBEAMS. For example, the access of the database using Python or
Java were based on the needs and the use of the languages for different
purposes. In this way, the implementation of the persistence model captures
the instances of any known data collected from NetBEAMS and saves it using the
document pattern as shown in Listing \ref{file:mongodb-ysi-data-format}. Any
programming language used was capable of accessing the same collected data by
their own.

In the end, the implementation of the data persistence for NetBEAMS was
developed in both Single and Distributed Systems. The use of regular single
server execution provided enough computational power for a compatible volume of
data when compared to the one produced from the SF-BEAMS sensor network. In
this way, the implementation suggested by this work fulfill the need of a
persistence system for a Small volume of data being created at the same type
in a Single Server. Furthermore, mongoDB also provided support for Distributed
System, when the use of databse shards, in forms of different mongoDB nodes,
were used as a clustered environment.

@@

>>>>

As discussed in the previous chapter, the implementation of a data persistence
layer for NetBEAMS followed the requirements defined in Chapter 4 in regards
to the classification of the SF-BEAMS sensor network and the software
infrastructure provided by the Data Sensor Platform. The implementation was
developed using the infrastructure provided by the online version-control
system repository from NetBEAMS, whose documentation is detailed in Section
\ref{sec:dsp-data-persistence-implementation}.

This chapter details the experiment setup designed to evaluate the
proposed implementation, showing the measurement results collected from the
simulation log and based on the user experience while reusing the collected
data. The pertinent discussion of this work is divided up into different
sections, each of them analyzing different aspects related to data persistence
for NetBEAMS, among others.

@@

The experiment was designed to support different scenarios of regular use of
the mongoDB system infrastructure, simulating the activities of data collection
from the SF-BEAMS network using the NetBEAMS infrastructure. First, the setup
for the data collection of randomly generated data was implemented, which
directly uses already existing NetBEAMS classes. The CRUD scenarios were
implemented in different programming languages, from the data insertion
scenario using the Java Programming Language, to the remaining operations using
the Javascript Scripting language \cite{javascript} and the Python Programming
Language \cite{python}. Similarly, an open-source web application, developed
in the Ruby Programming language \cite{ruby}, was also used. The experiment
scripts were written using Shell Bash Script Language \cite{bashshell} and is
responsible for orchestrating the execution of one single experimental ``run''.

In order to evaluate the DSP Data Persistence component, experiments were
conducted using the DSP Mouse Actions component, a software sensor simulator
that captures the actions performed by the mouse pointer.

@@

Different artifacts were created to perform the different experiments regarding
the CRUD operations. The ``Create'' operation' was developed in Java, isolating
the services provided by the class DSPMongoCRUDService (Listing
\ref{file:dsp-mongo-service}). It is responsible for the connection handling
with mongoDB, as well as the insertion operation from NetBEAMS. As described
\cite{netbeams-dsp-architecture}, the current version of NetBEAMS supports the
data collection for the YSI Sonde device \cite{YSI-Sonde}, among others used
for demo purposes. After evaluating the database with the YSI Sonde data
handler, the use of the demo component DSP Mouse Actions was used. It captures
the mouse pointer activities performed over a window, for example, mouse clicks
and mouse grag over coordinates x and y.

In order to exercise mongoDB with the same data collected from NetBEAMS, the
existing Java class TestSondeData (Listing
\ref{file:random-ysi-data-generator}), was refactored to provide a random
instance generator for the YSI Data handler implemented in the Java class
SondeDataType (Listing \ref{file:main-ysi-data-handler}). When the Insert
function is executed, the created instances are transferred from the JVM to
the mongoDB, and stored in the collection ``SondeDataContainer''. Specific
details about the execution of mongoDB are depicted in Section
\ref{sec:mongodb-deployment}. Finally, the implementation of the experiment
in the Java class DSPMessageToMongoDBExperiment (Listing
\ref{file:experiment-dsp-java-executor}) was responsible for managing the
creation and insertion of any workload composed by instances of the class
SondeDataType in a bulk manner.

@@

On the other hand, the execution of the main experiment shell script requires
the user to provide the size of the workload, as shown in listing
\ref{cmd:run-persistence-experiment}.

@@

The workload selected for the experiments reflects the current use of the
collected data at the SF-BEAMS infrastructure. In this way, the experiment
runs for 5 different times to investigate different scenarios. This number is
related to the total number of sensor devices reported to be in operation, as
discussed in Section \ref{sec:sfbeams}. The volume of the workload prepared
after the experiment setup is as follows:

\begin{itemize}
  \item \textbf{First Round}: 483,840 documents, worth of production of the 
  YSI Sonde device during one year, at the rate of 1 observation per minute;
  \item \textbf{Next Rounds}: 483,840 * 4, or 2,419,200 documents, representing
  5 different YSI devices similar to the first round.
\end{itemize}

@@

Two different system environments were prepared: one for the External Storage
environment, or Single Server; and another for the Data-Centric Storage, or
Distributed Server. Since the single server is a regular centralized database
system, mongoDB just requires the execution of the database process ``mongod'',
which is started by specifying the directory where the inserted data will be
managed. The distributed server setup requires a different list of processes
(see Section \ref{sec:mongodb-user-experience}): one ``mongod'' process for
each mongoDB shard running on the different mongoDB cluster, one ``mongod''
process to manage the cluster metadata, and one ``mongos'' process, which is
the main one responsible to orchestrate the cluster. The single system
environment is symmetric to the External Storage used to persist the collected
data, while the clustered one simulates the Data-Centric Storage one.

@@

In order to verify the feasibility of the solution proposed, all CRUD
operations were exercised through the implementation of the practical
scenarios based on the use cases described in Section \ref{sec:use-cases}. In
brief, the use cases were chosen to run with random values such as follows:

\begin{itemize}
  \item (R0) Insert a volume of collected data from sensors using NetBEAMS
  service compatible with 1 year of observations;
  \item (R1) Find all documents whose observations were collected between two
  different dates, say the days between December 8 and 15 of 2009;
  \item (R2) Find all documents whose observations were collected from a given
  known sensor device such as one whose IP address is equals to "192.168.0.102";
  \item (R3) Find all documents whose observations contained specific values
  read from the environment, say Salinity equals to 0.01 and the water temperature
  equals to 46.47;
  \item (U1) Update all the documents whose observations were collected during
  the day of December 2nd, 2009, by adding a tag = ``oil spill'';
  \item (D1) Remove all the documents whose observations were collected on
  December 7th, 2009.
  \item (R4) Export the produced data of the week to CSV files, using the same
  tabular format sequence used in the files distributed by SF-BEAMS using the 
  the OPeNDAP format, as in Listing \ref{file:rtc-ysi-opendap}.
\end{itemize}

@@

As described in the previous section, the design of the experiments were based
on the specifications of a data persistence for NetBEAMS, using the mongoDB
database system. After setting up the experiment environment, the execution of
the workload was performed against the mongoDB database using its mechanisms
used to manipulate the data generated. It uses the programming language
abstraction to provide access and modifiy the data collected from the NetBEAMS
component. For this reason, this section details these mechanisms used to
implement each of the scenarios defined in the previous section, associating
numerical response time collected from the logs.

@@

The first set of measurements collected were the ones regarding the use case
R1, shown in Listing \ref{file:dsp-mongo-service}, defined as the insertion of
instances of the collected data from the sensor devices using Java. The
document follows the specification of the keys designed in Section
\ref{sec:dsp-persistence-data-model}. The construction of the keys
and respective values were done using a Hash-like object ``BasicDBObject''
such as the ones shown in the method ``buildKeySegment''. Similarly, the
creation of database indexes was performed using the same approach, as shown in
the Java method ``setupMetadataIndexes''. Then, the insertion of the
documents was performed on the service method call
``insertPersistentUnitMessageContents'', which delimits the bulk creation using
the transaction management mechanism over the collection. In this way,
different insert times were collected from the logs of the different
experiments as summarized on Table 7.1.

\begin{table}[!h]
    \begin{center}
        \begin{tabular}{|p{100pt}|p{100pt}|p{100pt}|}\hline
           x & \textbf{Linux VirtualBox Server} & \textbf{Mac OS Host Server}\\\hline 
           \textbf{Single Server} & ~25K docs/sec & ~51K/sec \\\hline
           \textbf{Clustered Server} & ~17K docs/sec & ~39K/sec \\\hline
        \end{tabular}
        \caption{Insertion averages on Virtual and Host in Single or Clustered
        mongoDB Server}
    \end{center}
    \label{tab:experiment-insert-avarage}
\end{table}

After executing the experiment, the claimed disk space for the representation
of the randomly generated data was separated into two different sizes: one
related to the amount of data produced by one YSI device, and another produced
by five different YSI devices, as summarized on Table 7.2.

\begin{table}[!h]
    \begin{center}
        \begin{tabular}{|p{100pt}|p{100pt}|p{100pt}|}\hline
        x & \textbf{1 YSI} & \textbf{5 YSIs}\\\hline
        \textbf{Indexed Keys} & 278.33 MB & 1.35 GB \\\hline
        \end{tabular}
        \caption{Amount of disk storage used}
    \end{center}
    \label{tab:experiment-insert-diskspace}
\end{table}

@@

The simulation of the use case scenarios for the implementation used the mongoDB
shell client, which uses Javascript, for most of the operations for
``retrieve'', ``update'' and ``delete'' shown in Listing
\ref{file:experiment-scenarios}. The implementation of the use case to export
the collected data to OPeDAP format used Python, depicted on Listing
\ref{file:experiment-export-python}. The access pattern to use the collected
data is ``db.collection\underline{ }name.operation", where
``collection\underline{ }name'' is the name of the collection used for the
collected data, and in this case, the name of the DSP component that
represents the collected data. Therefore, mongoDB uses dynamic object binding
for the collection object, and executes the CRUD operations as method calls
with a list of the needed parameters. Table 7.3 relates method calls used to
implement each of the use cases defined, as implemented in Listing
\ref{file:experiment-scenarios}.

\begin{table}[!h]
    \begin{center}
        \begin{tabular}{|c|c|c|c|}\hline
        \textbf{Scenario Type} & \textbf{Language} & \textbf{Method Call} & \textbf{Use Case}\\\hline 
        \textbf{Insert} & Java & db.SondeDataContainer.insert() & R0\\\hline
        \textbf{Retrive} & Javascript & db.SondeDataContainer.find() & R1, R2, R3\\\hline 
        \textbf{Update} & Javascript & db.SondeDataContainer.update() & U1\\\hline 
        \textbf{Delete} & Javascript & db.SondeDataContainer.remove() & D1\\\hline
        \textbf{Export} & Python & db.SondeDataContainer.find() & R4\\\hline
        \end{tabular}
        \caption{mongoDB method calls used for the use cases}
    \end{center}
    \label{tab:experiment-scenarios-mongo-methods}
\end{table}

The execution time for each of the use cases were collected using the logs
collected during the execution of the experiment script manager. The retrieval
operations ran in average for 300ms, while the updates took more than 10
seconds when it involved larger data sets with more than 300,000 documents.
Similarly, the delete operation took less than 13 seconds to perform.

@@

In general, the user experience of mongoDB is related the methods used to
access and modify the state of the database using a different approach when
compared to traditional Relational databases. Since most of the methods to
create, access and modify the database were performed using the programming
language abstraction of objects and method calls, the transition from the use
of a relational data model is very smooth. The creation of data was the first
contact with mongoDB in terms of data creation in Java. Later, different
methods of data access were used after the randomly generated data was
inserted into mongoDB: using the shell commands mongo and ``mongoexport'',
command-line execution developed in Python, through the REST Web Services API,
and using a web application, written in the Ruby Programming Language
\cite{ruby}, providing the view over the collected data using a web browser.
More details about the execution of these commands during the experiments are
shown in Section \ref{sec:mongodb-user-experience}.

The primary method used to access the mongoDB database is using the shell
command. Once the user requests access to the database, he or she can issue an
operation over the collection of documents using regular the abstraction
programming language, as the shown in Listing \ref{file:experiment-scenarios}.
The CRUD operations are available to the user and can be used at any time to
any database collection. Aggregation operations such as to count the returned
result set is also provided via programming languages abstraction.

In order to export the collected data to other data formats, the  command
``mongoexport'' and the developed export script written in Python were used.
The ``mongoexport'', exports the collected data from the database using the
native support to comma-delimited values. As shown in Listing
\ref{file:mongodb-export-command}, the actual execution describe how mongo is
querying the data from the database. As a result, the total number of documents
exported is shown in on line 8. In addition to the regular command that
exports the entire database, the user can provide a query string during the
execution of the command in order to filter the dataset returned and,
consequently, the number of documents exported. Similar to the use of the
default export capabilities of mongoDB, a script written in Python to export
data from mongoDB was written as shown in Listing
\ref{file:experiment-export-python}. It uses the package ``pymongo'' to create
a comma-delimited file with the same header and line format as shown in Listing
\ref{file:experiment-export-python-results}. The results can be compared with
the example the output of the HTTP Response body of Listing
\ref{file:rtc-ysi-opendap}, requested directly to the OPeNDAP server of RTC.
Therefore, the experiment is capable of reproducing theat the OPeNDAP can.

While the access to the database using the shell command gives full access to 
the CRUD operations, the access to the data was also performed using the REST
Web Services API, as shown in Listing \ref{cmd:mongo-rest-request}. As one can
see, the IP address and the default port number of the mongoDB REST middleware
are provided. Then, the REST operation uses the name of the database and
the collection to correlate with the ones in the database system. Thus,
querying the collections from mongoDB can be performed using the HTTP Request
Parameter variables, by adding the prefix ``filter\underline{ }'' before the
name of the key of the document. The given example shows the query performed
in the collected data for the key ``observation.Conductivity'' with the value
``104.5''. Similarly, the HTTP Request variable ``limit'' is used to reduce
the number of documents returned by the query. The HTTP Response from mongoDB
returns the result using the JSON format, similar to the ones printed in the
command-shell.

@@

The actual execution of the solution was performed using the demo Mouse
Actions, which consists of a software sensor that perceives any mouse action in
a window area. The experiment consisted of moving the mouse around a white
window, where each action is captured and temporarily saved until a threshold is
reached, making the component to send the messages to the persistence layer, so
that it can be saved on mongoDB. Figure \ref{fig:experiment-mouseactions-dsp}
shows a screenshot of the moment when the message is sent to mongoDB, being
verified through the HTTP REST Interface using a web browser
\ref{sec:dsp-mongodb-rest-ws}.

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.4]{../diagrams/experiment-mouseactions-dsp}
  \caption{Experiment showing the mongoDB persisting messages from the DSP
  Mouse Actions Demo}
  \label{fig:experiment-mouseactions-dsp}
\end{figure}

@@

The implementation of the data persistence for NetBEAMS follows the
specifications explained in Chapter 4. The purpose of the sensor network for
NetBEAMS is to support data archival from any collected data, and for this
reason, mongoDB was selected after an empirical analysis described in Chapter
5. The requirements described in Chapter 4 were expressed as examples of use
cases in Section \ref{sec:exp-scenarios}. This chapter discusses the
experiments from different angles: analysis of the taxonomies, the
infrastructure, the use cases evaluation and applicability in other projects.

@@

Any database system covers the purpose of the sensor data for NetBEAMS, which
is Data Archival, along with the type of storage for the data as an External
Storage. After selecting mongoDB, the development of the persistence layer met
the requirements of archiving data for different types of sensor devices from
the DSP Platform such as the YSI Data Handler and the Mouse Actions demo data.
As it is shown in Section \ref{sec:exp-measurements}, the collected data is
archived in files using the BSON format \cite{bson}, a binary representation
used to store instances of the document described in Listing
\ref{file:mongodb-ysi-data-format}. Along with the purpose of Data Archival,
the use of mongoDB also met the specification of External Storage, since the
data persisted in the file system, as well as Data-Centric for running mongoDB
with different  distributed mongo shards (cluster nodes).

SF-BEAMS uses Tabular Data model, a schema-less data model that represents the
collected data using comma-delimited files, without a database system to manage
those files, just the OPeNDAP middleware. In contrast, the data persistence
method proposed by this work suggests mongoDB as a schema-less database system
to manage the data collected using NetBEAMS. As described in Chapter 5, the
schema-less data model can provide better support to Data Provenance because
the data model allows the change of records without affecting the
others. In this way, the metadata used by the implementation clearly describes
the three fundamental questions regarding the collected data from sensor
devices, as described in Section \ref{sec:dsp-persistence-data-model}. The 
identity of the data is given by the key ``message\underline{ }id'', in a way
that it uniquely identifies the data produced by the sensor identified by the
key ``sensor.ip\underline{ }address''. Thus, the message can be tracked
back to the sensor that produced the data for different purposes such as
tracking the data producer. Then, time dimensions were used in order to
track two different aspects of the data collection: the valid time can be used
to correlate observations to the exact moment it occurred, while the
transaction time can be used to verify when the data was collected, and make
decisions such as deleting the instances of data collected in the previous two
days, etc. Similarly, the use of the key ``sensor.location'' can give the exact
location of the data, since researchers can be interested in finding data
based on a specific location. The most important aspect of the schema-less data
model is that the observed data is added to the key ``observation'', as any
sensor device carries its set of attributes of key-value pairs. A good example
of the applicability of the schema-less model is the scalability regarding
changes to the data structure. For example, if an existing sensor device is
flushed with a new version of the firmware\footnote{Primary machine
instructions that run in the hardware}, changing the format of data types or
adding new values, the following instances of collected data may very
from the previous one by simply adding the new additions of data. For this
reason, the DSP Data Persistence gives administrators the ability to choose
which fields to persist by declaring it using a bootstrap message for the
component, as shown in the parameter ``YSI\underline{ }DESIRED\underline{
}PROPERTIES'' of Listing \ref{file:dsp-data-persistence-bootstrap.xml}.

In order to access the collected data, the implementation used different APIs
provided by the mongoDB community. The Centralized Query Mechanism was
satisfied by the use of those APIs, which also fulfills the requirements
defined by NetBEAMS. For example, the accesses of the database using Python or
Java were based on the needs and the use of the languages for different
purposes. In this way, the implementation of the persistence model captures
the instances of any known data collected from NetBEAMS and saves it using the
document pattern as shown in Listing \ref{file:mongodb-ysi-data-format}. Any
programming language used was capable of accessing the same collected data by
their own.

In the end, the implementation of the data persistence for NetBEAMS was
developed in both Single and Distributed Systems. The use of regular single
server execution provided enough computational power for a compatible volume of
data when compared to the one produced from the SF-BEAMS sensor network. The
implementation suggested by this work fulfill the needs of a persistence system
for a Small volume of data being created at the same time in a Single Server.
Furthermore, mongoDB also provided support for a Distributed System, when the
use of database shards, in forms of different mongoDB nodes, was used as a 
clustered environment.

@@

\subsection{Applicability and Recommendations}
